---
title: "Fitting Network Scale-up Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{networkscaleup-fitting-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
--

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This packaged fits several different network scale-up models (NSUM) to Aggregated Relational Data (ARD). ARD represents survey responses about how many people each respondents knows in different subpopulations through "How many X's do you know?" questions. Specifically, if $N_i$ respondents are asked how many people they know in $N_k$ subpopulations, then ARD is an $N_i$ by $N_k$ matrix, where the $(i,j)$ element represents how many people respondent $i$ reports knowing in subpopulation $j$. NSUM leverages these responses to estimate the unknown size of hard-to-reach populations. See Laga, et al. (2021) for a more details.

In this package, we provide functions to estimate the size and accompanying parameters (e.g. degrees) from 5 papers:
* Killworth et al. (1998a) plug-in MLE
* Killworth et al. (1998b) MLE
* Zheng et al. (2006) overdispersed model
* Teo et al. (2019) no covariate, barrier, and transmission models
* Laga et al. (2022+) uncorrelated, correlated, and covariate models

This vignette shows how to fit all of these models on one data set.

## Instructions

First, load the library:

```{r setup}
library(networkscaleup)
```

We will simulate a data set from the Binomial formulation in Killworth et al. (1998a).

```{r simulation}
set.seed(1998)
N_i = 500
N_k = 5
N = 1e5
sizes = rbinom(N_k, N, prob = runif(N_k, min = 0.01, max = 0.15))
degrees = round(exp(rnorm(N_i, mean = 5, sd = 1)))

ard = matrix(NA, nrow = N_i, ncol = N_k)
for(k in 1:N_k){
  ard[,k] = rbinom(N_i, degrees, sizes[k] / N)
}
```

### PIMLE

```{r pimle}

```
