---
title: "Networkscaleup: An R Package for Fitting and Evaluating Models 
for Aggregated Relational Data"
date: "2026-02-23"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Owen G. Ward
    affiliation: Simon Fraser University
    address:
    - Department of Statistics and Actuarial Science
    - Burnaby, BC, Canada
    url: https://owenward.github.io/
    orcid: 0000-0002-9678-3542
    email:  oward@sfu.ca
  - name: Benjamin Vogel
    affiliation: Montana State University
    address:
    - Department of Mathematical Sciences
    - Bozeman, MT, USA
    email: 
  - name: Jieyun Wang
    affiliation:
    - University of Kentucky
    address:
    - Dr. Bing Zhang Department of Statistics
    - Lexington, KY, USA
    email:
  - name: Anna L. Smith
    affiliation:
    - University of Kentucky
    address:
    - Dr. Bing Zhang Department of Statistics
    - Lexington, KY, USA
    url: https://annalantzsmith.github.io/
    email: asm297@g.uky.edu
    orcid: 0000-0003-1507-0207
  - name: Le Bao
    affiliation:
    - Pennsylvania State University
    address:
    - Department of Statistics
    - University Park, PA, USA
    url: https://lebao0215.github.io/
    email: lebao@psu.edu
    orcid: 0000-0003-0191-5918
  - name: Xiaoyue Niu
    affiliation:
    - Pennsylvania State University
    address:
    - Department of Statistics
    - University Park, PA, USA
    url: https://sites.google.com/view/maggiexniu/
    email: xiaoyue@psu.edu
    orcid: 0000-0003-3384-7446
  - name: Ian Laga
    affiliation:
    - Montana State University
    address:
    - Department of Mathematical Sciences
    - Bozeman, MT, USA
    url: https://ilaga.github.io/
    email: ian.laga@montata.edu
    orcid: 0000-0002-5164-4856
type: package
output: 
  rjtools::rjournal_pdf_article:
    toc: no
    header-includes:
      - \usepackage{bm}
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE,
                      message = FALSE, fig.height = 4)
library(networkscaleup)
library(ggplot2)
set.seed(100)

```

# Introduction

- ARD is ....
- ARD is widely used in ....

- In this paper we will review how to use \CRANpkg{networkscaleup} to fit and diagnose
popular models for ARD. 

- Aim to provide a suite of functions for applied practitioners who
wish to use the models for their own ARD, and understand how 
to compare and contrast popular models.


## Outline of the article
In Section \@ref(sec:background) we first review ARD and popular models for 
this data. In Section \@ref(sec:package) we introduce the
\CRANpkg{networkscaleup} package which provides a 
suite of functions to simulate ARD. 
Section \ref{sec:fitting} describes how popular ARD models
can be fit using this package
while Section \ref{sec:eval} discusses the diagnostic functions
available for assessing the fit of common ARD models to data.

# Background {#sec:background}

ARD is a popular tool to study social networks and estimate the 
size of hard-to-reach populations in such networks.
Since it was first considered [@bernard1991estimating], it has been
used to estimate the size of a range of hidden populations
[@bernard2010counting, @salganik2011assessing] and estimate
personal network size [@mccormick2010many]. Here we provide a brief overview
of ARD and the most commonly used methods, particularly the 
Network Scale-Up Method (NSUM). Detailed reviews of NSUM and general models
for ARD, along with important
implementation considerations, 
are given in @laga2021thirty, @mccormick2020network, @ward2025bayesian.


An ARD sample $Y$ takes the form of an $n \times K$ matrix, where 
individual entries $y_{ik}$ denote the number of people respondent $i$ knows
in subpopulation $k$. This data is typically collected by asking a sample
from a social network questions of the form "How many [X]'s do you know?",
where X represents each of the $K$ subpopulations for which data is collected.
These can represent people in the network with specific traits or
characteristics of interest, such as hard to reach populations (e.g, people
who use drugs).

ARD models seek to place distributional assumptions on the
entries of $Y$, such that properties of the underlying network
and respondents can be inferred.
The original and most widely used statistical model for ARD is the 
classic NSUM procedure [@killworth1998estimation, @killworth1998social],
which assumes that

\begin{equation}
Y_{ik} \sim \text{Binomial}(\exp(\alpha_i), \exp(\beta_k)),
(\#eq:simp-pois)
\end{equation} 


with $\alpha_i$ the log-degree of respondent $i$ and $\beta_k$
the log-prevalence of subpopulation $k$ in the underlying social network.
This initial model does not allow for the incorporation of covariates,
which are commonly collected and utilised in the models considered below.
<!-- Maybe a bit more discussion of this model here? -->

Two types of covariates have been considered in ARD models: respondent-level
covariates and respondent–group-level covariates. Respondent-level covariates are 
commonly collected, whereas respondent–group-level covariates are 
relatively rare.
In this work, we focus exclusively on respondent-level covariates and leave the
incorporation of respondent–group-level covariates in software to future work.
Let $Z$ denote the $n \times p$ matrix of respondent-level covariates, 
where the $i$-th row $z_i$ contains the covariates for respondent $i$. 
These covariates do not vary across groups indexed by $k$. 
Their effects on the response $Y_{ik}$ may be either global or local. 
A global covariate induces a shift in the expected value of $Y_{ik}$ 
that is constant across all groups $k$, given by a common coefficient $\beta$. 
In contrast, a local covariate has an effect that varies with $k$, 
allowing the impact of $z_i$ to differ across groups through a varying
coefficient $\beta_k$.
In the models described below, this difference corresponds to
whether the coefficient in $\beta^\mathsf{T}z_i$ varies
for different values of $k$.


Several extensions of this initial Binomial model have been 
considered which allow the incorporation of covariate information.
A natural extension is the Poisson model of @teo2019estimating, where
counts are modeled, with respondent covariates $z_i$, as

\begin{equation}
Y_{ik} \sim \text{Poisson}(\exp(\alpha_i +\beta_k + \beta^\mathsf{T}z_i)).
(\#eq:pois-cov)
\end{equation}


Alternatively, @zheng2006many utilised a Negative Binomial distribution
(without covariates)
for respondent counts, which allows group specific overdispersion through the
parameter $\omega_k$,

\begin{equation}
Y_{ik} \sim \text{Negative-Binomial}(\mu = \exp(\alpha_i +\beta_k + \beta^\mathsf{T}z_i),\ \text{overdispersion}=\omega_k).
\end{equation}

These models are widely considered in the ARD literature and serve as 
a set of initial candidate models for many ARD analyses. 
We note also that there have been several important extensions
of these models
in recent years, including incorporating correlation across 
groups [@laga2023correlated] and between degree and 
group [@vogel2025accounting].
There are also other related ARD models, which we do not consider here.
These include robust estimation for NSUM models [@diaz2026robust] and
dichotomous ARD models [@baum2023uses]. Similarly, we do not consider
ARD models for inferring the underlying network [e.g, @breza2020using]
or design based models [e.g, @feehan2016generalizing].


Given the variety of ARD models proposed in the literature, a key practical 
challenge is selecting an appropriate specification for a given dataset. This 
includes determining which covariates to include and which
distribution to assume 
for the outcome $Y$. The \CRANpkg{networkscaleup} package
provides a selection of functions to fit some of the
most widely used candidate models and
evaluate their fit for given data. In the next section, we
introduce the \CRANpkg{networkscaleup} package and 
describe how it can be used for evaluating and 
comparing ARD models.



# The networkscaleup package {#sec:package}


<!-- TO DO: Talk very briefly about installing the package here... -->

The \CRANpkg{networkscaleup} package provides a complete suite of 
functions to fit and do model checking for 
the most widely used ARD models.
It allows simulation of synthetic ARD from the widely used models described 
above, with the possibility to incorporate both local and/or
global covariates also, as is often the case in real data.
The researcher can then utilise either this simulated data or their
own data to fit a range of candidate models and identify 
what underlying structure is captured by a model. This provides users with
a principled way to expand the models they consider for their problem,
incorporating additional structure only when it is identified by the 
diagnostic functions described below.

We will illustrate the workflow possible using this package
with both simulated and real data.
We will 
first generate two synthetic datasets from the widely used Poisson
model described above in Equation\@ref(eq:simp-pois).
In the first case we will 
consider a simulated dataset where there are no covariates, meaning 
the underlying Poisson rates are given exactly by 

$$
Y_{ik}\sim \text{Poisson}(\exp(\alpha_i + \beta_k)).
$$
This is one of the simplest ARD models and 
data from this model can then easily be simulated, specifying the size
of the ARD sample and the underlying population.
To simulate data from this model, with no covariates, the 
function `make_ard` can be used, which can simulate ARD from either
a Poisson on Negative Binomial model, with arguments specifying the size
of the ARD sample $Y$ along with potential covariate structure
and population degree parameters. The following demonstrates a
minimal example of this function, generating a small ARD sample.

```{r sim_pois_data_1, echo =TRUE, eval=FALSE}
library(networkscaleup)
make_ard(family = "poisson", n_i = 500, n_k = 20, N = 1e06)
```

The underlying values of $\alpha_i$ are drawn from a normal distribution
(which can be specified) while the $\beta_k$ correspond to 
subpopulations drawn from a uniform distribution giving a size
between 1% and 15% of the total population
(a typical size in real ARD).


We can also incorporate respondent level covariates into this
model, by specifying the number of non-zero coefficients of each.
For example, the following code will generate $p=6$ covariates,
of which two have non-zero local coefficients and one 
will have a non-zero global coefficient.
The underlying coefficients and the covariates will then also be stored 
in the returned object, along with the ARD sample. This generates
more realistic synthetic data which we will use throughout this paper
to demonstrate the code functionality.

```{r sim_pois_data_2, echo = TRUE}
sim_dat <- make_ard(n_i = 500, n_k = 20, N = 1e06,
                    p = 6,
                    p_local_nonzero = 2,
                    p_global_nonzero = 1,
                    family = "poisson")
```

For example, here we have specified that only one of the 
global covariates has a non-zero coefficient, which is seen
in the following output.

```{r see-global-beta, echo=TRUE}
sim_dat$x_beta_global 
```

To illustrate the functionality of the \CRANpkg{networkscaleup}
with real data, we consider ARD collected in Ukraine in 2008-2009
[@paniotto2009estimating] which describes ARD for more than 9000 respondents
about 15 groups. Several groups were removed using the
pre-processing procedure outlined in @laga2023correlated. There
are 8 respondent level covariates here, including age, gender, employment
status and several additional binary education covariates.
The subpopulations respondents were queried about include
demographic categories (e.g., how many males aged 20-30 do you know?),
specific names (e.g., how many
people named Pavlo do you know?)
and specific activities (e.g., how many intravenous drug users do you know?).
This is a large ARD sample but it is representative of the 
data size collected in current ARD studies. 
Fitting complex ARD models to this data can be computationally
intensive. As such, being able to identify the appropriate form
of ARD model for such data is an important initial step which
can prevent fitting complex and expensive models which are 
not appropriate for a given dataset.


<!-- TO DO: Spruce this up a little. Maybe add some more background or -->
<!-- something -->


# Fitting models to ARD {#sec:fitting}

Given that we have either simulated ARD or have collected data which
is to be analysed, the first task
is to fit candidate models to this data, 
so that the fit of these models can be
assessed. \CRANpkg{networkscaleup} provides two main options for 
fitting these models.
The first uses \CRANpkg{glmmTMB} [@glmmtmb1, @glmmtmb2] to compute maximum 
likelihood estimates (MLEs) for each of the models, providing point estimates
of the underlying parameters. 
The second approach uses \CRANpkg{rstan} [@rstan],
a software for Bayesian inference which allows for the incorporation
of priors into the model. Here, we use Stan only for optimization, 
which also returns MAP (maximum a posteriori) point estimates.
The goal of both of these model fitting procedures is to provide the 
user with reasonable model estimates quickly, allowing them to 
identify plausible models. For large ARD, fitting a fully Bayesian
model can require considerable computation time. 
These functions
allow the user to quickly identify an appropriate model,
which can then be fit using MCMC methods, if required.
We use only the MLE estimates throughout
this paper, but similar results are generally obtained with MAP estimates.
As the goal of these functions is to obtain quick diagnostics of potential
models, either approach can be considered.

We first fit the standard Poisson model, ignoring covariate information,
to our simulated dataset using the function `fit_mle`. 
This function takes several arguments, including the ARD, any potential covariates
to include in the model, and the distributional family assumed 
for the model.
This returns appropriate parameter 
estimates (for $\alpha$ and $\beta$ in a model with no covariates), along
with Pearson residuals [@mccullagh1989generalized]
and Randomized Quantile residuals [@dunn1996randomized].
We obtain similar results using either residual but show only Randomized
Quantile residuals in this paper.

We can fit a standard Poisson model to 

```{r fit_pois_data, echo = TRUE}
pois_fit <- fit_mle(sim_dat$ard, family = "poisson")

pois_fit$alphas[1:5] # the first 5 log degree estimates
```


Similarly, we can fit the initial Poisson model of Equation~\@ref(eq:simp-pois)
to the Ukraine ARD. We initially ignore the covariates, before then
using our diagnostics to identify which covariates to incorporate.
This simple model fits quickly for this large dataset, giving MLE estimates
in less than 8 seconds on a recent laptop. We will then use the 
fit of this model to identify local and global covariates.

```{r}
y <- read.csv("paper/data/Ukraine_y_analysis.csv")
x <- read.csv("paper/data/Ukraine_dem_cov_analysis.csv")
cov_names <- colnames(x)
y_m <- as.matrix(y)
```

```{r fit_pois_ukr, echo=TRUE}
pois_fit_ukr <- fit_mle(ard = y_m, family = "poisson")
```

In each case this fitted object can then be used for subsequent 
diagnostics, which we describe below.


# Evaluating ARD models {#sec:eval}

Given the fit of a specified ARD model, the next goal is to identify 
the appropriateness of this fitted model and what structure (if any)
the model fails to capture. \CRANpkg{networkscaleup} provides a range 
of tools to assess this, including to diagnose specific forms
of model misspecification. 
The workflow we recommend for this task, for any fitted model, is
described in detail in [@laga2026evaluating]. In brief, the recommended
procedure is:

1. Identify appropriate covariate structure. 
1. Test for correlation patterns across groups.
1. Evaluate distributional assumptions.


We will describe the functions available in \CRANpkg{networkscaleup}
for each of these steps, along with considering them for 
our simulated and real data examples.

## Identifying Covariate Structure

To identify which covariates should be considered
for inclusion in a potential model, the
`cov_plots` function examines the relationship between these covariates
and the fitted model,
which can then be used to identify which local and global covariates should
be included. 
This function generates two plots. The first, shown in 
Figure \@ref(fig:plot-cov-poisson) for our simulated example, 
plots a fitted smooth relationship (the default is a linear regression)
along with associated confidence bounds, between the values of the covariates
and their corresponding 
residuals, for each of the subpopulations in the ARD.
A relationship between these covariates and residuals can be used to
identify potential local covariates.


```{r make-plot-cov-poisson, echo = TRUE, fig.cap="Local covariates", fig.pos='H'}
cov_plot <- cov_plots(
  ard = sim_dat$ard,
  model_fit = pois_fit,
  x_cov = sim_dat$x_cov,
  se = T
)

local_cov_plot <- cov_plot$Group_plot
global_cov_plot <- cov_plot$Respondent_plot
```


```{r plot-cov-poisson, echo = TRUE, fig.cap="Local covariates", fig.height=3, fig.pos='H'}
local_cov_plot 
```

The second plot, shown in Figure \@ref(fig:plot-global-cov),
displays the smooth relationship between the covariate values
and the estimated log-degree for each respondent, 
for each of the potential covariates.
Linear patterns in these figures indicates covariates which 
are correlated with the estimated respondent degree, which can 
then be incorporated as global covariates.


<!-- The resulting figures,  -->
<!-- for our simulated data with covariates, -->
<!-- are shown  -->
<!-- in Figure \@ref(fig:plot-cov-poisson) and Figure \@ref(fig:plot-global-cov). -->
<!-- We see that for this simulation example, these plots correctly identify -->
<!-- the true local and global coefficients in this data. -->



```{r plot-global-cov, echo=TRUE, fig.cap="Global covariates", fig.height=2.5, fig.pos='H'}
global_cov_plot 
```

For this simulated data, this plots clearly identify the true local and global 
covariates. Given the result of this diagnostic, we refine our
model for the simulated data,
now incorporating the local
and global covariates 
suggested by Figures \@ref(fig:plot-cov-poisson),\@ref(fig:plot-global-cov).

```{r fit-cov-sim, echo=TRUE}
pois_fit_cov <- fit_mle(
  sim_dat$ard,
  x_cov_local = as.matrix(sim_dat$x_cov[, c(1, 3)]),
  x_cov_global = as.matrix(sim_dat$x_cov[, c(2)]),
  family = "poisson"
)
```

If we repeat this diagnostic after incorporating these covariates
we see either very weak or no relationship in both these covariate
diagnostics.

We repeat this analysis for our Ukraine ARD. We first examine the 
local covariates which are shown in Figure~\@ref(fig:plot-local-cov-ukr).
Here, Age, Gender, Employed, Access to Internet and Secondary Education
as showing linear patterns, indicating they should be chosen for
inclusion in any potential model.

```{r get-cov-ukraine}
cov_pois <- cov_plots(
  ard = y_m,
  model_fit = pois_fit_ukr,
  x_cov = x,
  se = T
)

local_cov_plot_ukr <- cov_pois$Group_plot
global_cov_plot_ukr <- cov_pois$Respondent_plot
```

```{r plot-local-cov-ukr, echo=TRUE, fig.cap="Local covariates, Ukraine", fig.height=5, fig.pos='H'}
local_cov_plot_ukr 
```


Similarly, we plot global covariates in Figure~\@ref(fig:plot-global-cov-ukr).
Here we show all covariates but only identify those not chosen as
local covariates. This identifies the 3 remaining 
education covariates as potential global effects.

```{r plot-global-cov-ukr, echo=TRUE, fig.cap="Global covariates, Ukraine", fig.height=2.5, fig.pos='H'}
global_cov_plot_ukr
```


Given this identification of local and global covariates, 
we can then refine our fitted model to incorporate this specification.
Incorporating these covariates increases the computational 
time required to fit this model, from several seconds to several minutes.

```{r fit-cov-ukr, echo=TRUE}
pois_fit_ukr_cov <- fit_mle(
  y_m,
  x_cov_local = as.matrix(x[, c(1, 2, 4, 5, 6)]),
  x_cov_global = as.matrix(x[, c(3, 7, 8)]),
  family = "poisson"
)
```



## Diagnosing Group Correlation

Next we wish to test for correlation across groups. 
[@laga2026evaluating] describe an eigenvalue based test for this,
where under the null hypothesis (of no group correlation) the test
statistic should follow a Tracy-Widom distribution [@tracy1996orthogonal].
We show these test statistics and the null distribution in 
Figure \@ref(fig:plot-tw-sim) for our second simulated dataset, 
after refitting the model to include the covariates suggested by 
our previous diagnostics. This function, `tw_group_corr_test`, also
returns additional information which is used in constructing this test,
and a corresponding p-value.
Note that incorporating these covariates gives a
plausible test statistic, indicating that, after accounting 
for these covariates, there is limited evidence for
any remaining group correlation
structure.

```{r plot-tw-sim, fig.cap="TW Test",fig.pos='H', echo=TRUE, fig.height=2.5, fig.width=5}
tw_test <- tw_group_corr_test(model_fit = pois_fit_cov, plot = TRUE)
tw_test$p_value
tw_test$tw_density_plot
```



```{r plot-tw-ukr, fig.cap="TW Test, Ukraine Data", fig.pos='H',fig.height=2.5, fig.width=5, echo=TRUE}
tw_ukr <- tw_group_corr_test(model_fit = pois_fit_ukr_cov, plot = TRUE)
tw_ukr$p_value
tw_ukr$tw_density_plot
```



## Checking Distributional Assumptions


Given we have investigated the presence of correlation across groups,
we finally wish to test the distributional assumptions underlying our 
model. This can be done using the function
`hang_rootogram_ard`
to construct a hanging rootogram for a
fitted model [@kleiber2016visualizing].
In Figure \@ref(fig:root) we show this diagnostic
for the model fit to our simulated data which incorporates the
true covariate structure. We see that the rootogram indicates good model fit,
with all bars close to the x-axis, indicating agreement 
between the expected and observed frequencies.
We also have a dispersion test, which is shown for this data in 
Figure \@ref(fig:disp-sim). These dispersion tests are significant 
for one of the twenty subpopulations, as would be expected by chance
at $\alpha = 0.05$.

```{r root, echo = TRUE, fig.height=2.5, fig.cap="Hanging rootogram for simulated data."}
pois_root <- hang_rootogram_ard(
  ard = sim_dat$ard,
  model_fit = pois_fit_cov,
  x_max = 100
) 
pois_root

```


```{r disp-sim, fig.height=2.5, fig.cap="Dispersion test for simulated data across subpopulations"}
disp_test <- dispersion_metric(ard = sim_dat$ard, model_fit = pois_fit_cov)
disp_test$plots
```


We can repeat these distributional tests for our real data.

```{r root-ukr, fig.height=3.5, fig.cap="Overall Rootogram for Ukraine Data"}
pois_root_ukr <- hang_rootogram_ard(
  ard = y_m,
  model_fit = pois_fit_ukr_cov
)
pois_root_ukr
```


```{r fig.height=2.5, fig.cap="Dispersion test for Ukraine data across subpopulations"}
disp_test_ukr <- dispersion_metric(ard = y_m, model_fit = pois_fit_ukr_cov)
disp_test_ukr$plots
```


## Additional Diagnostic Functions

While we have described the key workflow for constructing an 
appropriate ARD model for data, \CRANpkg{networkscaleup}
also contains some additional general diagnostic functions which may 
provide additional insights.
The function `plot_fitted` takes the residuals from
a fitted ARD model and plots them against the expected values of $Y_{ik}$
under that model, as is commonly done to assess model residuals.
In Figure \@ref(fig:fitted-no-cov) we show this plot with Randomized Quantile
residuals for our simulation with covariates, where we use the fitted model
ignoring these covariates.
Figure \@ref(fig:fitted-cov) shows the residuals when we incorporate
the residuals chosen by our diagnostic tests. We see that this indicates
improved model fit, as expected.


```{r fitted-no-cov, echo=TRUE, fig.height=2.5, fig.cap="Plotting fitted residuals indicates lack of model fit when covariates are ignored."}

plot_fitted(ard = sim_dat$ard,
            model_fit = pois_fit,
            resid = "rqr") +
  labs(title = "Ignoring covariates")
```



```{r fitted-cov, echo=TRUE, fig.height=2.5, fig.cap="Incorporating covariates indicates improved model fit."}
plot_fitted(ard = sim_dat$ard,
            model_fit = pois_fit_cov,
            resid = "rqr") +
  labs(title = "Including covariates")
```


```{r ukraine-cov, echo=TRUE, fig.height=2.5, fig.cap="Plotting fitted residuals indicates lack of model fit, after including covariates."}
plot_fitted(ard = y_m,
            model_fit = pois_fit_ukr_cov,
            resid = "rqr") +
  labs(title = "Ignoring covariates")

```



# Summary
- A summary of what we have done, and why we have done it
