---
title: "Networkscaleup: An R Package for Fitting and Evaluating Models 
for Aggregated Relational Data"
date: "2026-01-19"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Quietest Quokka
    affiliation: University of Little Mates
    address:
    - Department of Letter Q
    - Somewhere, Australia
    url: https://www.britannica.com/animal/quokka
    orcid: 0000-1721-1511-1101
    email:  qquo@ulm.edu
  - name: Bounciest Bilby
    affiliation:
    - University of Little Mates
    - University of Aussie Animals
    address:
    - Department of Letter Q, Somewhere, Australia
    - Department of Marsupials, Somewhere, Australia
    url: https://www.britannica.com/animal/bilby
    email: bbil@ulm.edu
    orcid: 0000-0002-0912-0225
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(networkscaleup)
library(palmerpenguins)
library(ggplot2)
set.seed(100)

```

# Introduction

- ARD is ....
- ARD is widely used in ....

- In this paper we will review how to use \CRANpkg{networkscaleup} to fit and diagnose
popular models for ARD. 

- Aim to provide a suite of functions for applied practitioners who
wish to use the models for their own ARD, and understand how 
to compare and contrast popular models.


## Outline of the article



# Background

- A brief but self contained description of ARD and the network scale up method in general. 

- Then this will be followed by the goal of collecting widely used models
for this data, and providing software to fit these models quickly 
(fit and fail fast). 


# The networkscaleup package

The \CRANpkg{networkscaleup} package provides a complete suite of 
functions to fit and do model checking for 
the most widely used models for ARD.
It allows simulation of synthetic ARD from the widely used models described 
above, with the possibility to incorporate both local and/or
global covariates also, as is often the case in real data.

The researcher can then utilise either this simulated data or their
own data to fit a range of candidate models and identify 
what underlying structure is captured by a model. This provides users with
a principled way to expand the models they consider for their problem,
incorporating additional structure only when it is identified by the 
diagnostic functions described below.

To illustrate the workflow possible using this package, we will 
first generate two synthetic datasets from the widely used Poisson
model described above (reference eqn). In the first case we will 
consider a simulated dataset where there are no covariates, meaning 
the underlying Poisson rates are given exactly by 

$$
Y_{ik}\sim \text{Poisson}(\exp(\alpha_i + \beta_k)).
$$

Data from this model can then easily be simulated, specifying the size
of the ARD sample and the underlying population.

```{r sim_pois_data_1, echo =TRUE}
sim_dat_1 <- make_ard(family = "poisson", n_i = 500, n_k = 20, N = 1e06)
```

The underlying values of $\alpha_i$ are drawn from a normal distribution
(which can be specified) while the $\beta_k$ correspond to 
uniform subpopulations of size between 1% and 15% of the total population
(a typical size in real ARD).



We can also incorporate respondent and group level covariates into this
model, by specifying the number of non-zero coefficients of each.
For example, the following code will generate $p=6$ covariates,
of which two have non-zero local coefficients and one 
will have a non-zero group coefficient.
The underlying coefficients and the covariates will then also be stored 
in this object.

```{r sim_pois_data_2, echo = TRUE}
sim_dat_2 <- make_ard(p = 6,
                      p_local_nonzero = 2,
                      p_global_nonzero = 1,
                      family = "poisson")

sim_dat_2$x_beta_global ## to see the single non-zero global coefficient
```


TO DO: Incorporate a real data example here also, describe how to read the 
data in and store it so that it can be used like the simulated data 
in everything below.


# Fitting Models to ARD

<!-- - Describe the key functionality we have to do this -->

<!-- - A worked example with simulated data from the models in the package -->

<!-- - A worked example with real data, where we fit these models. -->

Given that we have either simulated ARD or used real data, the first task
is to fit models to this data, so that the fit of these models can then be
assessed. \CRANpkg{networkscaleup} provides two main options for fitting these models.
The first uses \CRANpkg{glmmTMB} [@glmmtmb1] [@glmmtmb2] to compute maximum 
likelihood estimates (MLEs) for each of the models, providing point estimates
of the underlying parameters. The second approach uses \CRANpkg{rstan} [@rstan],
a software for Bayesian inference which allows for the incorporation
of priors into the model. Here, we use Stan only for optimization, 
which also returns (maximum a posteriori) point estimates.
The goal of both of these model fitting procedures is to provide the 
user with reasonable model estimates quickly, allowing them to 
identify plausible models. For large ARD, fitting a fully Bayesian
model can require considerable computation time. The goal of these functions
is to quickly identify an appropriate model, which can then be fit
using software to generate the underlying posterior density rather 
than just point estimates.

We first fit the standard Poisson model, ignoring covariate information,
to both simulated datasets using function `fit_mle`, which can
specify the ARD, the form of the ARD model to be fit, and 
covariates to include in the model also. This returns parameter 
estimates (for $\alpha$ and $\beta$ in this case), along
with Pearson residuals and Randomized Quantile residuals 
(cite these here).


```{r fit_pois_data, echo = TRUE}
pois_fit_1 <- fit_mle(sim_dat_1$ard, family = "poisson")
pois_fit_2 <- fit_mle(sim_dat_2$ard, family = "poisson")

pois_fit_1$alphas[1:10] # the log degree estimates
```



# Evaluating ARD Models



```{r plot_fitted_poisson, fig.cap="Test Caption"}
plot_fitted(
  ard = sim_dat_1$ard,
  model_fit = pois_fit_1,
  resid = "rqr"
)
# 
# plot_fitted(
#   ard = sim_dat_2$ard,
#   model_fit = pois_fit_2,
#   resid = "pearson"
# )
```



<!-- We also have a function to construct a hanging rootogram for these -->
<!-- fitted models (cite). -->

<!-- ```{r plot_root_poisson, echo = TRUE} -->
<!-- pois_root_1 <- hang_rootogram_ard( -->
<!--   ard = sim_dat_1$ard, -->
<!--   model_fit = pois_fit_1 -->
<!-- ) -->
<!-- pois_root_1 -->

<!-- pois_root_2 <- hang_rootogram_ard( -->
<!--   ard = sim_dat_2$ard, -->
<!--   model_fit = pois_fit_2, -->
<!--   x_max = 200 -->
<!-- ) -->
<!-- pois_root_2 -->
<!-- ``` -->


```{r plot_cov_poisson, echo = TRUE}
cov_p_2 <- cov_plots(
  ard = sim_dat_2$ard,
  model_fit = pois_fit_2,
  x_cov = sim_dat_2$x_cov,
  se = T
)

local_cov_plot <- cov_p_2$Group_plot
global_cov_plot <- cov_p_2$Respondent_plot

# local_cov_plot # Suggests x2 and x4 correctly
# global_cov_plot # Suggests x1 correctly
```




- Describe the key functions for this in the package

- Continue the two worked examples from before, illustrating the uses of the
package for these tasks specifically


# Summary

- A summary of what we have done, and why we have done it
