---
title: "Networkscaleup: An R Package for Fitting and Evaluating Models 
for Aggregated Relational Data"
date: "2026-02-23"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Owen G. Ward
    affiliation: Simon Fraser University
    address:
    - Department of Statistics and Actuarial Science
    - Burnaby, BC, Canada
    url: https://owenward.github.io/
    orcid: 0000-0002-9678-3542
    email:  oward@sfu.ca
  - name: Benjamin Vogel
    affiliation: Montana State University
    address:
    - Department of Mathematical Sciences
    - Bozeman, MT, USA
    email: 
  - name: Jieyun Wang
    affiliation:
    - University of Kentucky
    address:
    - Dr. Bing Zhang Department of Statistics
    - Lexington, KY, USA
    email:
  - name: Anna L. Smith
    affiliation:
    - University of Kentucky
    address:
    - Dr. Bing Zhang Department of Statistics
    - Lexington, KY, USA
    url: https://annalantzsmith.github.io/
    email: asm297@g.uky.edu
    orcid: 0000-0003-1507-0207
  - name: Ian Laga
    affiliation:
    - Montana State University
    address:
    - Department of Mathematical Sciences
    - Bozeman, MT, USA
    url: https://ilaga.github.io/
    email: ian.laga@montata.edu
    orcid: 0000-0002-5164-4856
type: package
output: 
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 4)
library(networkscaleup)
library(ggplot2)
library(patchwork)
set.seed(100)

```

# Introduction

- ARD is ....
- ARD is widely used in ....

- In this paper we will review how to use \CRANpkg{networkscaleup} to fit and diagnose
popular models for ARD. 

- Aim to provide a suite of functions for applied practitioners who
wish to use the models for their own ARD, and understand how 
to compare and contrast popular models.


## Outline of the article



# Background

ARD is a popular tool to study social networks and estimate the 
size of hard-to-reach populations in such networks.
Since it was first considered [@bernard1991estimating], it has been
used to estimate the size of a range of hidden populations
[@bernard2010counting] [@salganik2011assessing] and estimate
personal network size [@mccormick2010many]. Here we provide a brief overview
of ARD and the most commonly used methods, particularly the 
Network Scale-Up Method (NSUM). Detailed reviews of NSUM and general models
for ARD are given in @laga2021thirty, @mccormick2020network, @ward2025bayesian.


An ARD sample $Y$ takes the form of an $n \times K$ matrix, where 
individual entries $y_{ik}$ denote the number of people respondent $i$ knows
in subpopulation $k$. This data is typically collected by asking a sample
from a social network questions of the form "How many [X]'s do you know?",
where X represents each of the $K$ subpopulations for which data is collected.
These can represent people in the network with specific traits or
characteristics of interest, such as hard to reach populations (e.g, people
who use drugs).

The original and most widely used statistical model for ARD is the 
classic NSUM procedure [cite this], which assumes that

$$
Y_{ik} \sim \text{Binomial}(\exp(\alpha_i), \exp(\beta_k)),
$$ 

with $\alpha_i$ the log-degree of respondent $i$ and $\beta_k$
the log-prevalence of subpopulation $k$ in the underlying social network.
This initial model does not allow for the incorporation of covariates,
which are commonly collected and which we discuss below.


- A brief but self contained description of ARD and the network scale up method in general. 

- Then this will be followed by the goal of collecting widely used models
for this data, and providing software to fit these models quickly 
(fit and fail fast). 

Diagnostics paper talks about how negative binomial models 
don't include covariates so don't consider them further, just mention
them briefly here (same as that paper).


# The networkscaleup package

The \CRANpkg{networkscaleup} package provides a complete suite of 
functions to fit and do model checking for 
the most widely used models for ARD.
It allows simulation of synthetic ARD from the widely used models described 
above, with the possibility to incorporate both local and/or
global covariates also, as is often the case in real data.

The researcher can then utilise either this simulated data or their
own data to fit a range of candidate models and identify 
what underlying structure is captured by a model. This provides users with
a principled way to expand the models they consider for their problem,
incorporating additional structure only when it is identified by the 
diagnostic functions described below.

We will illustrate the workflow possible using this package
with both simulated and real data.
We will 
first generate two synthetic datasets from the widely used Poisson
model described above (reference eqn). In the first case we will 
consider a simulated dataset where there are no covariates, meaning 
the underlying Poisson rates are given exactly by 

$$
Y_{ik}\sim \text{Poisson}(\exp(\alpha_i + \beta_k)).
$$
This is one of the simplest ARD models and 
data from this model can then easily be simulated, specifying the size
of the ARD sample and the underlying population.

```{r sim_pois_data_1, echo =TRUE}
sim_dat_1 <- make_ard(family = "poisson", n_i = 500, n_k = 20, N = 1e06)
```

The underlying values of $\alpha_i$ are drawn from a normal distribution
(which can be specified) while the $\beta_k$ correspond to 
subpopulations drawn from a uniform distribution giving a size
between 1% and 15% of the total population
(a typical size in real ARD).



We can also incorporate respondent and group level covariates into this
model, by specifying the number of non-zero coefficients of each.
For example, the following code will generate $p=6$ covariates,
of which two have non-zero local coefficients and one 
will have a non-zero group coefficient.
The underlying coefficients and the covariates will then also be stored 
in this object.

```{r sim_pois_data_2, echo = TRUE}
sim_dat_2 <- make_ard(n_i = 500, n_k = 20, N = 1e06,
                    p = 6,
                    p_local_nonzero = 2,
                    p_global_nonzero = 1,
                    family = "poisson")

sim_dat_2$x_beta_global ## to see the single non-zero global coefficient
```


TO DO: Incorporate a real data example here also, describe how to read the 
data in and store it so that it can be used like the simulated data 
in everything below. Even though this is not particularly interesting.
Could just keep the simulated data to this section only...


# Fitting models to ARD

<!-- - Describe the key functionality we have to do this -->

<!-- - A worked example with simulated data from the models in the package -->

<!-- - A worked example with real data, where we fit these models. -->

Given that we have either simulated ARD or have collected data which
is to be analysed, the first task
is to fit candidate models to this data, 
so that the fit of these models can be
assessed. \CRANpkg{networkscaleup} provides two main options for 
fitting these models.
The first uses \CRANpkg{glmmTMB} [@glmmtmb1] [@glmmtmb2] to compute maximum 
likelihood estimates (MLEs) for each of the models, providing point estimates
of the underlying parameters. The second approach uses \CRANpkg{rstan} [@rstan],
a software for Bayesian inference which allows for the incorporation
of priors into the model. Here, we use Stan only for optimization, 
which also returns MAP (maximum a posteriori) point estimates.
The goal of both of these model fitting procedures is to provide the 
user with reasonable model estimates quickly, allowing them to 
identify plausible models. For large ARD, fitting a fully Bayesian
model can require considerable computation time. 
These functions
allow the user to quickly identify an appropriate model,
which can then be fit using MCMC methods, if required.
We use the MLE estimates for illustration
in this paper, but similar results are obtained with MAP estimates.

We first fit the standard Poisson model, ignoring covariate information,
to both simulated datasets using the function `fit_mle`. 
This function takes several arguments, including the ARD, any potential covariates
to include in the model, and the distributional family assumed 
for the model.
This returns appropriate parameter 
estimates (for $\alpha$ and $\beta$ in a model with no covariates), along
with Pearson residuals [@mccullagh1989generalized]
and Randomized Quantile residuals [@dunn1996randomized].
We obtain similar results using either residual but show only Randomized
Quantile residuals in this paper.


```{r fit_pois_data, echo = TRUE}
pois_fit_1 <- fit_mle(sim_dat_1$ard, family = "poisson")
pois_fit_2 <- fit_mle(sim_dat_2$ard, family = "poisson")

pois_fit_1$alphas[1:5] # the first 5 log degree estimates
```


TO DO: Fit the real data here also.


# Evaluating ARD models


Given the fit of a specified ARD model, the next goal is to identify 
the appropriateness of this fitted model and what structure (if any)
the model fails to capture. \CRANpkg{networkscaleup} provides a range 
of tools to assess this, including to diagnose specific forms
of model misspecification. 
The workflow we recommend for this task, for any fitted model, is
described in detail in [@laga2026evaluating]. In brief, the recommended
procedure is:

1. Identify appropriate covariate structure. 
1. Test for correlation patterns across groups.
1. Evaluate distributional assumptions.


We will describe the functions available in \CRANpkg{networkscaleup}
for each of these steps, along with considering them for 
each of our examples.

## Identifying Covariate Structure

To identify which covariates may be required to in the model, the
`cov_plots` function displays the relationship between these covariates,
which can then be used to identify which local and global covariates should
be included in the model. The resulting figures, 
for our simulated data with covariates,
are shown 
in Figure \@ref(fig:plot-cov-poisson) and Figure \@ref(fig:plot-global-cov).
We see that for this simulation example, these plots correctly identify
the true local and global coefficients in this data.


```{r plot-cov-poisson, echo = TRUE, fig.cap="Local covariates", fig.pos='H'}
cov_p_2 <- cov_plots(
  ard = sim_dat_2$ard,
  model_fit = pois_fit_2,
  x_cov = sim_dat_2$x_cov,
  se = T
)

local_cov_plot <- cov_p_2$Group_plot
global_cov_plot <- cov_p_2$Respondent_plot

local_cov_plot # Suggests x2 and x4 correctly
```

```{r plot-global-cov, echo=TRUE, fig.cap="Global covariates", fig.height=2.5, fig.pos='H'}
global_cov_plot # Suggests x1 correctly
```



TO DO: Repeat with the McCarthy Data here.

Given the result of these diagnostics, we refine our
model for the simulated data which does have covariates,
now incorporating the local
and global covariates 
suggested by Figures \@ref(fig:plot-cov-poisson),\@ref(fig:plot-global-cov).

```{r, echo=TRUE}
pois_fit_2_cov <- fit_mle(
  sim_dat_2$ard,
  x_cov_local = as.matrix(sim_dat_2$x_cov[, c(2, 4)]),
  x_cov_global = as.matrix(sim_dat_2$x_cov[, c(1)]),
  family = "poisson"
)
```


## Diagnosing Group Correlation

Next we wish to test for correlation across groups. 
[@laga2026evaluating] describe an eigenvalue based test for this,
where under the null hypothesis (of no group correlation) the test
statistic should follow a Tracy-Widom distribution [@tracy1996orthogonal].
We show these test statistics and the null distribution in 
Figure \@ref(fig:plot-tw-sim) for our second simulated dataset, 
after refitting the model to include the covariates suggested by 
our previous diagnostics. This function, `tw_group_corr_test`, also
returns additional information which is used in constructing this test,
and a corresponding p-value.
Note that incorporating these covariates gives a
plausible test statistic, indicating that, after accounting 
for these covariates, there is not any remaining group correlation
structure.

```{r plot-tw-sim, fig.cap="TW Test",fig.pos='H', echo=TRUE, fig.height=2.5, fig.width=5}
tw_test <- tw_group_corr_test(model_fit = pois_fit_2_cov, plot = TRUE)
tw_test$p_value
tw_test$tw_density_plot
```



TO DO: Repeat this for the McCarthy data here.


## Checking Distributional Assumptions


Given we have investigated the presence of correlation across groups,
we finally wish to test the distributional assumptions underlying our 
model. This can be done using the function
`hang_rootogram_ard`
to construct a hanging rootogram for a
fitted model [@kleiber2016visualizing].
In Figure \@ref(fig:root) we show this diagnostic
for the model fit to our simulated data which incorporates the
true covariate structure. We see that the rootogram indicates good model fit,
with all bars close to the x-axis, indicating agreement 
between the expected and observed frequencies.
We also have a dispersion test, which is shown for this data in 
Figure \@ref(fig:disp-sim). As expected, these dispersion tests are not 
significant for any subpopulations.

```{r root, echo = TRUE, fig.height=2.5, fig.cap="Hanging rootogram for simulated data."}
root_2 <- pois_root_2 <- hang_rootogram_ard(
  ard = sim_dat_2$ard,
  model_fit = pois_fit_2_cov,
  x_max = 200
) 
root_2

```


```{r disp-sim, fig.height=2.5, fig.cap="Dispersion test for simulated data across subpopulations"}
disp_test <- dispersion_metric(ard = sim_dat_2$ard, model_fit = pois_fit_2_cov)
disp_test$plots
```


TO DO: Repeat for the McCarthy data here.

## Additional Diagnostic Functions

While we have described the key workflow for constructing an 
appropriate ARD model for data, \CRANpkg{networkscaleup}
also contains some additional general diagnostic functions which may 
provide additional insights.
The function `plot_fitted` takes the residuals from
a fitted ARD model and plots them against the expected values of $Y_{ik}$
under that model, as is commonly done to assess model residuals.
In Figure \@ref(fig:fitted-no-cov) we show this plot with Randomized Quantile
residuals for our simulation with covariates, where we use the fitted model
ignoring these covariates.
Figure \@ref(fig:fitted-cov) shows the residuals when we incorporate
the residuals chosen by our diagnostic tests. We see that this indicates
improved model fit, as expected.


```{r fitted-no-cov, echo=TRUE, fig.height=2.5, fig.cap="Plotting fitted residuals indicates lack of model fit when covariates are ignored."}

plot_fitted(ard = sim_dat_2$ard,
            model_fit = pois_fit_2,
            resid = "rqr") +
  labs(title = "Ignoring covariates")
```



```{r fitted-cov, echo=TRUE, fig.height=2.5, fig.cap="Incorporating covariates indicates improved model fit."}
plot_fitted(ard = sim_dat_2$ard,
            model_fit = pois_fit_2_cov,
            resid = "rqr") +
  labs(title = "Including covariates")
```


TO DO: Repeat for the McCarthy data here also.






# Summary

- A summary of what we have done, and why we have done it
